Teaching Low-Level Concurrency with \pebbles and \landslide

### 2016-08-27 sigcse-17/410.tex
sigcse: last minute changes

The first project is a backtrace library: %``stack crawler'':
when invoked by a client program, it displays the
%program's
stack symbolically,
showing function names, strings, floating-point values, etc.
This project
%\shortversion{enables}{meets a variety of goals: it enables}
%students to review key process-model and
%language-runtime concepts from the
%prerequisite course;
%it
introduces course expectations about
design, analysis, and making choices;
%finally,
and
because C pointers are unsafe, it requires students
to consider robustness.
%
% and we READ THEIR CODE
% ^^ LOL, classic dave.
This project uses a standard C tool chain
% FIXME garth wants to cite gdb
and debugged as students see fit (generally GDB, a symbolic debugger, and \x{printf}).

@@

\begin{table}[t]
	\center \footnotesize
	\begin{tabular}{|l|p{0.30\textwidth}|}
		\hline
		\bf Syscall name & \bf Summary \\
		\hline
		\multicolumn{2}{c}{\em Lifecycle management} \\
		\hline
		\x{fork} & Duplicates the invoking task, including all memory regions. \\
		\x{thread_fork} & Creates a thread in the current task.\\
		\x{exec} & Replaces the current program in the invoking task with a new one. \\
		\x{set_status} & Records exit status of current task. \\
		\x{vanish} & Ends execution of the calling thread. \\
		\x{wait} & Blocks execution until another task terminates; collects its exit status.\\
		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
		\hline
		\multicolumn{2}{c}{\em Thread management} \\
		\hline
		\x{gettid} & Returns the ID of the invoking thread. \\
		\x{yield} & Defers execution to a specified thread. \\
		\x{deschedule} & Blocks execution of invoking thread. \\
		\x{make_runnable} & Wakes another \x{deschedule}d thread. \\
		\x{get_ticks} & Gets timer tick count since bootup. \\
		\x{sleep} & Blocks for a given number of ticks. \\
		\x{swexn} & Registers a user-space function as a software exception handler.\\
		\hline
		\multicolumn{2}{c}{\em Memory management} \\
		\hline
		\x{new_pages} & Allocates a specified memory region. \\
		\x{remove_pages} & Deallocates same. \\
		\hline
		\multicolumn{2}{c}{\em Console I/O} \\
		\hline
		\x{getchar}* & Reads one character from keyboard. \\
		\x{readline} & Reads the next line from keyboard. \\
		\x{print} & Prints a memory buffer to the console. \\
		\x{set_term_color} & Sets color for future console output. \\
		\x{set_cursor_pos} & Sets the console cursor location. \\
		\x{get_cursor_pos} & Retrieves the console cursor location. \\
		\hline
		\multicolumn{2}{c}{\em Miscellaneous} \\
		\hline
		\x{ls} & Loads a buffer with the names of files stored in the RAM disk ``file system.'' \\
		\x{halt} & Ceases execution of the OS. \\
		\x{misbehave}* & Selects thread-scheduling policy. \\
		\hline
	\end{tabular}
	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement calls marked with an asterisk (*), though the reference kernel provides them. }
	\label{tab:syscalls}
\end{table}

@@

{\bf Lifecycle.} The process lifecycle is loosely based on the
Unix \x{fork()}/\x{exec()}/\x{wait()} model.
We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
distinction between ``tasks,''
which are resource containers,
and threads,
each of which executes within a single task.
This differs from %partitioning is a different approach from that of
Plan~9's \x{rfork()}~\cite{Pike90plan9} or Linux's \x{clone()},
%both of
which allow a continuum of sharing
%between identical threads and independent processes,
at the cost of implementation complexity. %which we wish to avoid.
%A task's address space can be edited through two system
%calls that add and delete contiguous regions of memory.
%Students are expected to implement zero-fill-on-demand,
%but copy-on-write and demand-loading of executables are
%optional.
%Threads can print text to and read lines of text from
%the simple console that students implemented at
%the start of the semester.

%Widely regarded as
The most difficult concurrency problem in the project
is that of coordinating a parent and a child task that ``simultaneously''
exit:
% invoke \x{vanish()}:
when a task completes,
live children and exited zombies must be handed off
to the task's parent or to the system's \x{init} process,
at a time when the task's parent may itself be
exiting;
meanwhile, threads in tasks that receive new children
may need to be awakened from the \x{wait()} system call.
Due to design constraints imposed by other parts of the kernel specification,
careless solutions %solutions which are not carefully designed
are prone to data races or deadlocks.

@@

\textbf{32-bit x86.}
%\subsubsection{x86}
%\shortversion{}{Students write code that runs on one specific
%hardware platform.}
We target x86 PCs %because machines of this type are
%available from many manufacturers at many price points.
%This means that it is easy for
so students can easily obtain access
to a machine (virtual or real) which will let them
test their code.
%boot
%up their code, type commands, and observe the results.
%\violence{}{Also, unlike many System-on-a-Chip platforms,
%detailed hardware descriptions are readily available.}
%% rather than being proprietary or "documented"
%% by labyrinthine Linux source code
%\shortversion{}{Finally, the platform is fairly stable---with only a
%little care, it is possible to write a compliant
%kernel which runs on anything newer than a 486DX processor,
%i.e., almost any PC hardware which is likely to be in
%working condition.}
%
%\subsubsection{32-bit}
We target the 32-bit
subset of x86, for multiple reasons.
%%% I'm not sure I believe this is an advantage. Instead of having
%%% to verify if each register is correct, students now have to verify
%%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
%First, while the small number of registers is
%frustrating for compiler writers,
%\shortversion{the}{from the point of view of OS students, the}
%existence of fewer registers means less work
%when it is necessary to verify that each register's
%value is correct.
%Likewise,
The calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater
insight.
%\shortversion{insight.}{insight. Finally, since it is impossible for an x86 processor
%to be in 64-bit mode with virtual memory off,
%students would need to be coached through the
%32-bit--to--64-bit transition or would need
%to be given a template virtual-memory implementation.}
%All in all, at present we are satisifed with
%the 32-bit choice.

>>>>

The first project is a backtrace library: %``stack crawler'':
when invoked by a client program, it displays the
%program's
stack symbolically,
showing function names, strings, floating-point values, etc.
This project
%\shortversion{enables}{meets a variety of goals: it enables}
%students to review key process-model and
%language-runtime concepts from the
%prerequisite course;
%it
introduces course expectations about
design, analysis, and making choices;
%finally,
and
because C pointers are unsafe, it requires students
to consider robustness.
%
% and we READ THEIR CODE
% ^^ LOL, classic dave.
This project uses a standard C tool chain
% FIXME garth wants to cite gdb
and debugged as students see fit (generally GDB, a symbolic debugger; and \x{printf}).

@@

\begin{table}[t]
	\center \footnotesize
	\begin{tabular}{|l|p{0.30\textwidth}|}
		\hline
		\bf Syscall name & \bf Summary \\
		\hline
		\multicolumn{2}{c}{\em Lifecycle management} \\
		\hline
		\x{fork} & Duplicates the invoking task, including all memory regions. \\
		\x{thread_fork} & Creates a thread in the current task.\\
		\x{exec} & Replaces the current program in the invoking task with a new one. \\
		\x{set_status} & Records exit status of current task. \\
		\x{vanish} & Ends execution of the calling thread. \\
		\x{wait} & Blocks execution until another task terminates; collects its exit status.\\
		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
		\hline
		\multicolumn{2}{c}{\em Thread management} \\
		\hline
		\x{gettid} & Returns the ID of the invoking thread. \\
		\x{yield} & Defers execution to a specified thread. \\
		\x{deschedule} & Blocks execution of invoking thread. \\
		\x{make_runnable} & Wakes another \x{deschedule}d thread. \\
		\x{get_ticks} & Gets timer tick count since bootup. \\
		\x{sleep} & Blocks for a given number of ticks. \\
		\x{swexn} & Registers a user-space function as a software exception handler.\\
		\hline
		\multicolumn{2}{c}{\em Memory management} \\
		\hline
		\x{new_pages} & Allocates a specified memory region. \\
		\x{remove_pages} & Deallocates same. \\
		\hline
		\multicolumn{2}{c}{\em Console I/O} \\
		\hline
		\x{getchar}* & Reads one character from keyboard. \\
		\x{readline} & Reads the next line from keyboard. \\
		\x{print} & Prints a memory buffer to the console. \\
		\x{set_term_color} & Sets color for future console output. \\
		\x{set_cursor_pos} & Sets the console cursor location. \\
		\x{get_cursor_pos} & Retrieves same. \\ %the console cursor location. \\
		\hline
		\multicolumn{2}{c}{\em Miscellaneous} \\
		\hline
		\x{ls} & Loads a buffer with the names of files stored in the RAM disk ``file system.'' \\
		\x{halt} & Ceases execution of the OS. \\
		\x{misbehave}* & Selects thread-scheduling policy. \\
		\hline
	\end{tabular}
	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement calls marked with an asterisk (*), though the reference kernel provides them. }
	\label{tab:syscalls}
\end{table}

@@

{\bf Lifecycle.} The process lifecycle is loosely based on the
Unix \x{fork()}/\x{exec()}/\x{wait()} model.
We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
distinction between {\em tasks},
which are resource containers,
and {\em threads},
each of which executes within a single task.
This differs from %partitioning is a different approach from that of
Plan~9's \x{rfork()}~\cite{Pike90plan9} or Linux's \x{clone()},
%both of
which allow a continuum of sharing
%between identical threads and independent processes,
at the cost of implementation complexity. %which we wish to avoid.
%A task's address space can be edited through two system
%calls that add and delete contiguous regions of memory.
%Students are expected to implement zero-fill-on-demand,
%but copy-on-write and demand-loading of executables are
%optional.
%Threads can print text to and read lines of text from
%the simple console that students implemented at
%the start of the semester.

%Widely regarded as
The most difficult concurrency problem in the project
is that of coordinating a parent and a child task that simultaneously
exit:
% invoke \x{vanish()}:
when a task completes,
live children and exited zombies must be handed off
to the task's parent or to the system's \x{init} process,
at a time when the task's parent may itself be
exiting;
meanwhile, threads in tasks that receive new children
may need to be awakened from the \x{wait()} system call.
Due to design constraints imposed by other parts of the kernel specification,
careless solutions %solutions which are not carefully designed
are prone to data races or deadlocks.

@@

\textbf{32-bit x86.}
%\subsubsection{x86}
%\shortversion{}{Students write code that runs on one specific
%hardware platform.}
We target x86 PCs %because machines of this type are
%available from many manufacturers at many price points.
%This means that it is easy for
so students can easily obtain access
to a machine (virtual or real) which will let them
test their code.
%boot
%up their code, type commands, and observe the results.
%\violence{}{Also, unlike many System-on-a-Chip platforms,
%detailed hardware descriptions are readily available.}
%% rather than being proprietary or "documented"
%% by labyrinthine Linux source code
%\shortversion{}{Finally, the platform is fairly stable---with only a
%little care, it is possible to write a compliant
%kernel which runs on anything newer than a 486DX processor,
%i.e., almost any PC hardware which is likely to be in
%working condition.}
%
%\subsubsection{32-bit}
We target the 32-bit
subset of x86 for multiple reasons.
%%% I'm not sure I believe this is an advantage. Instead of having
%%% to verify if each register is correct, students now have to verify
%%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
%First, while the small number of registers is
%frustrating for compiler writers,
%\shortversion{the}{from the point of view of OS students, the}
%existence of fewer registers means less work
%when it is necessary to verify that each register's
%value is correct.
%Likewise,
The calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting any greater
insight.
%\shortversion{insight.}{insight. Finally, since it is impossible for an x86 processor
%to be in 64-bit mode with virtual memory off,
%students would need to be coached through the
%32-bit--to--64-bit transition or would need
%to be given a template virtual-memory implementation.}
%All in all, at present we are satisifed with
%the 32-bit choice.
